Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations

### 2021-07-13 submission/manuscript.rmd
draft to be shared with co-authors

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level may be
accurate, these estimated do not apply to individual patients, especially for
patients with predicted risk at the boundaries of the risk intervals. Hence, the
risk-stratified approach is useful for exploring and presenting HTE, but is not
useful for supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [cite PATH !!]. We aimed to summarize and compare different
risk-based models for predicting individualized treatment effects. We simulated
RCT settings to compare the performance of these models under different
assumptions of the relationship between baseline risk and treatment. We
illustrated the different models by a case study of predicting individualized
effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
patients with an acute myocardial infarction (MI).

@@

For each patient we generated 8 baseline covariates here
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
was allocated using a 50:50 split. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates. In
the base case scenario coefficient values were such, that the AUC of the
logistic regression model was $0.75$ and the event rate in the control arm was
$20\%$ (Supplement 1).

While binary outcomes in the control arm were generated from Bernoulli variables
with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
the outcomes in the treatment arm based on true probabilities 
$\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$

The coefficients $\gamma_0,\gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario to mimic a wide
variety scenarios, ranging from true constant relative treatment effect
($\gamma_1=\gamma_2=0$) to moderate and strong linear ($\gamma_2=0$) and
quadratic deviations. We also considered scenarios with treatment-covariate
interactions.
These scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

The sample size of the base case scenario was set to 4,250 ($80%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,064 and 17,000, respectively. We also evaluated the of worse
or better discriminative ability for risk, adjusting the baseline covariate
coefficients, such that the AUC of the regression model in the control arm
was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simualation study of 66 scenarios
(exact settings in the supplementary material).

@@

Second, we considered a model which assumes *constant relative treatment effect* (constant OR)
Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

>>>>

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level may be
accurate, these estimates do not apply to individual patients, especially for
patients with predicted risk at the boundaries of the risk intervals. Hence, the
risk-stratified approach is useful for exploring and presenting HTE, but is not
useful for supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
different risk-based models for predicting individualized treatment effects. We
simulated RCT settings to compare the performance of these models under
different assumptions of the relationship between baseline risk and treatment.
We illustrated the different models by a case study of predicting individualized
effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
patients with an acute myocardial infarction (MI).

@@

For each patient we generated 8 baseline covariates here
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
was allocated using a 50:50 split. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates. In
the base case scenario coefficient values were such, that the AUC of the
logistic regression model was $0.75$ and the event rate in the control arm was
$20\%$.

While binary outcomes in the control arm were generated from Bernoulli variables
with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
the outcomes in the treatment arm were based on true probabilities 
$\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$

The coefficients $\gamma_0,\gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario to mimic a wide
variety scenarios, ranging from true constant relative treatment effect
($\gamma_1=\gamma_2=0$) to moderate and strong linear ($\gamma_2=0$) and
quadratic deviations. We also considered scenarios with treatment-covariate
interactions.
These scenarios included 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions.

The sample size of the base case scenario was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,064 and 17,000, respectively. We also evaluated the effect of
worse or better discriminative ability for risk, adjusting the baseline
covariate coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 66 scenarios
(exact settings in the supplementary material).

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant OR). Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

### 2021-07-13 submission/manuscript.tex
draft to be shared with co-authors

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. However, even though estimates at
the risk subgroup level may be accurate, these estimated do not apply to
individual patients, especially for patients with predicted risk at the
boundaries of the risk intervals. Hence, the risk-stratified approach is
useful for exploring and presenting HTE, but is not useful for
supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested
various risk-based models including a prognostic index of baseline risk
(PI) and treatment assignment {[}cite PATH !!{]}. We aimed to summarize
and compare different risk-based models for predicting individualized
treatment effects. We simulated RCT settings to compare the performance
of these models under different assumptions of the relationship between
baseline risk and treatment. We illustrated the different models by a
case study of predicting individualized effects of tissue plasminogen
activator (tPA) versus streptokinase treatment in patients with an acute
myocardial infarction (MI).

@@

For each patient we generated 8 baseline covariates here
\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Treatment was allocated using a 50:50 split. Outcomes for patients in
the control arm were generated from a logistic regression model
including all baseline covariates. In the base case scenario coefficient
values were such, that the AUC of the logistic regression model was
\(0.75\) and the event rate in the control arm was \(20\%\) (Supplement
1).

While binary outcomes in the control arm were generated from Bernoulli
variables with true probabilities
\(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
the treatment arm based on true probabilities \(\text{expit}(lp_1)\),
with \[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]

The coefficients \(\gamma_0,\gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario to mimic
a wide variety scenarios, ranging from true constant relative treatment
effect (\(\gamma_1=\gamma_2=0\)) to moderate and strong linear
(\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
with treatment-covariate interactions. These scenarios include 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of \(66\) scenarios. The exact settings
for each scenario are available in the supplementary material.

The sample size of the base case scenario was set to 4,250 (\(80%
\) power for the detection of a marginal OR of 0.8). We evaluated the
effect of smaller or larger sample sizes of 1,064 and 17,000,
respectively. We also evaluated the of worse or better discriminative
ability for risk, adjusting the baseline covariate coefficients, such
that the AUC of the regression model in the control arm was 0.65 and
0.85 respectively.

Combining all these settings resulted in a simualation study of 66
scenarios (exact settings in the supplementary material).

@@

Second, we considered a model which assumes \emph{constant relative
treatment effect} (constant OR) Hence, absolute benefit is predicted
from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).

>>>>

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. However, even though estimates at
the risk subgroup level may be accurate, these estimates do not apply to
individual patients, especially for patients with predicted risk at the
boundaries of the risk intervals. Hence, the risk-stratified approach is
useful for exploring and presenting HTE, but is not useful for
supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested
various risk-based models including a prognostic index of baseline risk
(PI) and treatment assignment {[}3,4{]}. We aimed to summarize and
compare different risk-based models for predicting individualized
treatment effects. We simulated RCT settings to compare the performance
of these models under different assumptions of the relationship between
baseline risk and treatment. We illustrated the different models by a
case study of predicting individualized effects of tissue plasminogen
activator (tPA) versus streptokinase treatment in patients with an acute
myocardial infarction (MI).

@@

For each patient we generated 8 baseline covariates here
\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Treatment was allocated using a 50:50 split. Outcomes for patients in
the control arm were generated from a logistic regression model
including all baseline covariates. In the base case scenario coefficient
values were such, that the AUC of the logistic regression model was
\(0.75\) and the event rate in the control arm was \(20\%\).

While binary outcomes in the control arm were generated from Bernoulli
variables with true probabilities
\(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
the treatment arm were based on true probabilities
\(\text{expit}(lp_1)\), with
\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]

The coefficients \(\gamma_0,\gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario to mimic
a wide variety scenarios, ranging from true constant relative treatment
effect (\(\gamma_1=\gamma_2=0\)) to moderate and strong linear
(\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
with treatment-covariate interactions. These scenarios included 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions.

The sample size of the base case scenario was set to 4,250 (\(80\%\)
power for the detection of a marginal OR of 0.8). We evaluated the
effect of smaller or larger sample sizes of 1,064 and 17,000,
respectively. We also evaluated the effect of worse or better
discriminative ability for risk, adjusting the baseline covariate
coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 66
scenarios (exact settings in the supplementary material).

@@

Second, we considered a model which assumes \emph{constant relative
treatment effect} (constant OR). Hence, absolute benefit is predicted
from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).
